---
title: "real_data_analysis"
author: "Anne STAFF, Bianca BOI"
date: "2023-05-01"
output: pdf_document
---

## Real data analysis

Preparation (dependencies & data):
```{r}
library(fitdistrplus)
gpigs <- read.table("surv.gpigs.txt", header = T, sep = ";")
gpigs.noncensored <- gpigs[gpigs$censored == 0,]

gpigs.mc <- gpigs.noncensored$lifetime[gpigs.noncensored$regime == "M_C"]
gpigs.mc.with_censored <- gpigs$lifetime[gpigs$regime == "M_C"]
gpigs.m43 <- gpigs.noncensored$lifetime[gpigs.noncensored$regime == "M_4.3"]
```

### Description of the data

Provide meaningful descriptive statistics to compare the two treatments M4.3 and M.C.

```{r}
summary(gpigs.mc.with_censored)
```

```{r}
summary(gpigs.m43)
```

### Distribution fit

Using the R-package 'fitdistrplus', fit a distribution to the 'lifetime' variable for both control and treatment taking into account the right censoring. Start with Exponential distribution. Interpret the parameter estimates.

#### Exponential distribution

```{r warning=FALSE}
fe <- fitdist(gpigs.mc, "exp")
fe$estimate
```
TO DO: interpret the rate estimate

```{r}
denscomp(list(fe))
```
TO DO: describe why it's not adapted


#### Weibull, Gamma, log-Normal fit on the control group

```{r warning=FALSE}
fw <- fitdist(gpigs.mc, "weibull")
fg <- fitdist(gpigs.mc, "gamma") 
fln <- fitdist(gpigs.mc, "lnorm") 
plot.legend <- c("Weibull", "lognormal", "gamma")
denscomp(list(fw, fln, fg), legendtext = plot.legend) 
qqcomp(list(fw, fln, fg), legendtext = plot.legend) 
cdfcomp(list(fw, fln, fg), legendtext = plot.legend) 
ppcomp(list(fw, fln, fg), legendtext = plot.legend)
```

#### Weibull, Gamma, log-Normal fit on the 4.3 group

```{r warning=FALSE}
fw <- fitdist(gpigs.m43, "weibull")
fg <- fitdist(gpigs.m43, "gamma") 
fln <- fitdist(gpigs.m43, "lnorm") 
plot.legend <- c("Weibull", "lognormal", "gamma")
denscomp(list(fw, fln, fg), legendtext = plot.legend) 
qqcomp(list(fw, fln, fg), legendtext = plot.legend) 
cdfcomp(list(fw, fln, fg), legendtext = plot.legend) 
ppcomp(list(fw, fln, fg), legendtext = plot.legend)
```
```{r}
fw$estimate
```
TO DO : interpret estimates

```{r}
fg$estimate
```
TO DO : interpret estimates

```{r}
fln$estimate
```
TO DO : interpret estimates

#### log-Logistic
Define the distribution function & compagnie
```{r}
dloglogis <- function(x, alpha, beta) {
  res <- (beta/alpha) * (x/alpha)^(beta-1)*(1+(x/alpha)^beta)^-2
  return(res)
}

ploglogis <- function(q, alpha, beta) {
  res <- 1 / (1 + (q/alpha)^beta)
  return(res)
}

qloglogis <- function(p, alpha, beta) {
  res <- alpha * ((1/p) - 1)^(1/beta)
  return(res)
}
```

##### Control group
```{r warning=FALSE}
# MC regime (sans données censorées pour le dist)
fll <- fitdist(gpigs.mc, "loglogis", method = "mle", start=list(alpha=10, beta=5))
fll$estimate
denscomp(list(fll), main = "MC regime")
```

##### M4.3-Regime group
```{r warning=FALSE}
# M4.3 regime
fll <- fitdist(gpigs.noncensored$lifetime[gpigs.noncensored$regime == "M_4.3"], "loglogis", method = "mle", start=list(alpha=10, beta=5))
fll$estimate
```
TO DO: interpret estimates

```{r warning=FALSE}
denscomp(list(fll), main = "M4.3 regime")
```

